Malacology: A Programmable Storage System

### 2016-10-18 paper/paper.tex
paper updates
Signed-off-by: Noah Watkins <noahwatkins@gmail.com>

%These three trends put evolutionary pressure on storage systems and raise the
%question of whether 
The evolutionary pressure placed on storage systems by these trends raises the
question of whether
there are principles that storage systems designers can follow
to evolve storage systems efficiently, without jeopardizing years of
code-hardening and performance optimization efforts.
% that are important for
%users to continue to entrust their data to the storage system.

In this paper we investigate an approach that focuses on 
%generalizing 
%and
%composing 
identifying and exposing existing
storage system resources, services, and abstractions that in a generalized form
can be used to \emph{program} new services. By composing higher-level services
over existing abstraction one can reuse
subsystems and leverage their optimizations, established correctness,
robustness, and efficiency. 
%We will refer to this programmability as
%\emph{programmable storage}, which differs from \emph{active storage} (the
%injection and execution of arbitrary code in a storage system) 
%and
%\emph{software-defined storage} (the control of thin-provisioning of storage).

{\it \textbf{We define a programmable storage system}} to be a storage system that facilitates the 
re-use and extension of existing storage abstractions provided by the underlying software stack,
to enable the creation of new services via composition.
A programmable storage system can be realized by exposing existing functionality (such as 
metadata services and synchronization and monitoring capabilities) as interfaces that can be ``glued together''
in a variety of ways
using a high-level language.  Programmable storage differs from \emph{active storage},
or the injection and execution of arbitrary code in a storage system.

@@

Our prototype includes a non-exhaustive set of interfaces that can be used as building blocks
for constructing novel storage abstractions, including:

@@

Building storage stacks from the ground up for a specific purpose results in
the best performance. For example, HDFS was designed specifically to serve
Hadoop's jobs~\paa{lol not sure if I buy that hadoop achieves ``best performance''}, and uses techniques like exposing data locality and relaxing POSIX constraints to achieve
the best performance~\cite{CITEME}. Alternatively, general-purpose storage
stacks are built with the flexiblity to serve many applications by providing a
variety of interfaces and tunable parameters. Unfortunately, managing competing
forces in these systems is difficult and users
want more control from the general-purpose storage stacks
without going as far as building their storage system from the ground up.

@@

The Ceph community provides empirical evidence that developers are already beginning to embrace
programmable storage. Figure~\ref{fig:obj-int-dev-growth} shows a dramatic
growth in the production use of domain-specific interfaces in the Ceph
community since 2010. In that figure, classes are functional groupings of methods on storage
objects.  What is most remarkable is that this trend contradicts the notion
that API changes are a burden for users.  Rather it appears that a gap in
existing interfaces are being addressed through ad-hoc approaches to programmability. In fact,
Table~\ref{table:objclasses} categorizes existing interfaces and we clearly see
a trend towards reusable services.

@@

The takeaway from Figure~\ref{fig:obj-int-dev-growth} is that programmers are
already trying to use programmability because their needs, whether they be
related to performance, availability, consistency, convenience, etc., are not
satisfied by the existing default set of interfaces. The popularity of the custom object
interface facility of Ceph could be due to a number of reasons, such as the
default algorithms/tunables of the storage system are insufficient for the
application's performance goals, programmers want to exploit
application-specific semantics, and/or programmers know how to manage resources
to improve performance. Their solution is a way to work around the
traditionally rigid storage APIs because custom object interfaces give
programmers the ability to tell the storage system about their application: if
it is CPU or IO bound, if it has locality, if its size has the potential to
overload a single proxy node, etc.  The programmers know what the problem is
and how to solve it, but until the ability to modify object interfaces, they
had no way to express to the storage system how to handle their data.

Our approach is to expose more of the commonly used, code-hardened subsystems
of the underlying storage system as interfaces. The intent is that these
interfaces, which can be as simple as a redirection to the persistent data
store or as complicated as a strongly consistent directory service, should be
used and re-used in many contexts to implement a wide range of services. By
making programmability a `feature', rather than a `hack' or `workaround', we
help standardize a development process that is largely ad-hoc.

@@

Malacology implements and demonstrates the ideas of programmable storage in
Ceph. We choose Ceph because it is a production quality system and because it
is open source. The large developer community ensures that code is relatively
robust and the visibility of the code lets us expose any interface we want. In
the next section we describe the Ceph components that we expose as Malacology
interfaces.

@@

There are multiple reasons that make Lua an attractive runtime for implementing
Malacology. Lua is a portable, embedded scripting language that offers superior
performance and productivity trade-offs, including a JIT-based implementation
that is well known for near native performance. Additionally, Lua has been used
extensively in game engines, and systems research \cite{neto:dls14-luaos},
including storage systems where it has been effectively used both on
\cite{grawinkel:pdsw2012-lua,watkins2013:bdmc13-in-vivo,geambasu_comet_2010}
and off \cite{sevilla:sc15-mantle} the performance critical path. Finally, the
flexibility of the runtime allows to sandbox applications in order to address
security and performance concerns.

@@

In the case of Ceph, a consistent view of cluster state among server daemons
and clients is critical to provide strong consistency guarantees to clients.
Ceph maintains cluster state information in per-subsystem ``maps'' that record
membership and status information.  A PAXOS-based monitoring service is
responsible for integrating state changes into cluster maps, responding to
requests from out-of-date clients and synchronizing members of the cluster
whenever there is a change in a map so that they all observe the same system
state. As a fundamental building block of many system designs, consensus
abstractions such as PAXOS are a common technique for maintaining consistent
data versions, and are a useful system to expose.

The default behavior of the monitor can be seen as a PAXOS-based notification
system, similar to the one introduced in~\cite{burrows_chubby_2006}, allowing
clients to identify when new values (termed epochs in Ceph) are associated to
given maps. While Ceph does not expose this service directly, it does allow
limited use by clients through a key-value service intended for low-volume,
small size read-only configuration information. While a key-value service is
generally useful, it doesn't provide many of the useful services hidden within
the monitoring framework that would be required for applications with more
demanding requirements, such as creating arbitrary maps or associating
application-specific logic that can be executed for particular maps (or values
in a map).  Since the monitor is intended to be out of high-performance I/O
paths, a general guideline is to make use of this functionality infrequently
and to assign small values to maps.

@@

The second service implemented on Malacology is ZLog, a high-performance
distributed shared-log that is based on the CORFU
protocol~\cite{balakrishnan_corfu_2012}. A shared-log is a general, yet
powerful abstraction used to build many distributed systems, such as
cloud-based metadata management~\cite{balakrishnan:sosp13} and elastic database
storage engines~\cite{bernstein:cidr11,bernstein:vldb11,bernstein:sigmod15}.
However, existing implementations that rely on consensus algorithms such as
PAXOS funnel I/O through a single point introducing a bottleneck that restricts
throughput.  In contrast, the CORFU protocol is able to achieve high throughput
using a network counter service, called a \emph{sequencer}, for obtaining the
entry at the tail of log. This metadata operation that avoids all persistent
I/O in the common case.

@@

In order to repopulate the cached tail value during recovery of a sequencer,
the maximum position in the log must be obtained. To do this, the storage
interface exposes an additional \emph{seal} method that atomically installs a
new epoch value and returns the maximum log position that has been written.
Since the recovery protocol does not inform clients of a new epoch value until
the end of the process, the validity of the tail can be guaranteed despite the
iterative process of contacting storage devices.  Recovery of the sequencer
process itself may be handled in many ways, such as leader election using an
auxiliary service such as PAXOS. In our implementation, the recovery is the
same (and is inherited from) the Ceph metadata service.

@@

In Section~\ref{consistencyversioning-of-cluster-state} we described an
interface management service that is responsible for the durable and
consistent cluster-wide distribution of interface versions. We evaluate the
costs of such a service by embedding interface definitions in a Ceph cluster
membership data structure that is managed by a PAXOS instance, and distributed
using a scalable, peer-to-peer gossip protocol that is built into Ceph. We
assume an architecture that consists of a light-weight interface that enforces
the semantics of interface management resulting in one or more interface
updates that must be consistent across the cluster. The following experiments
focus on the cost of achieving distribution of consistent a interface version.

@@

Figure~\ref{fig:propdelay} shows the CDF of the latency of installing and
distributing an interface update on all OSDs in the cluster. The latency is
defined as the elapsed time following the PAXOS proposal for an interface
update until each OSD makes the update live (the cost of the PAXOS proposal is
configurable and is discussed below). The latency measurements were taken on
the nodes running Ceph server daemons, and thus exclude the client round-trip
cost. In each of the experiments 1000 interface updates were observed.

@@

\textbf{The Malacology Approach:} Re-use Ceph's Monitor functionality to
propagate application-specific state that needs to be highly-available and
strongly consistent.  The real cost of interface management, in addition to
cluster-wide propagation of interface updates, includes the network round-trip
to the interface management service, the PAXOS commit protocol, and other
factors such as system load. By default PAXOS proposals occur periodically with
a 1 second interval in order to accumulate updates. In a minimum, realistic
quorum of 3 monitors using HDD-based storage, we were able to decrease this
interval to an average of 222 ms.

>>>>

%These three trends put evolutionary pressure on storage systems and raise the
%question of whether 
The evolutionary pressure placed on storage systems by these trends raises the
question of whether
there are principles that storage systems designers can follow
to evolve storage systems efficiently, without jeopardizing years of
code-hardening and performance optimization efforts.
% that are important for
%users to continue to entrust their data to the storage system.
In this paper we investigate an approach that focuses on 
%generalizing 
%and
%composing 
identifying and exposing existing
storage system resources, services, and abstractions that in a generalized form
can be used to \emph{program} new services. By composing higher-level services
over existing abstraction one can reuse
subsystems and leverage their optimizations, established correctness,
robustness, and efficiency. 
%We will refer to this programmability as
%\emph{programmable storage}, which differs from \emph{active storage} (the
%injection and execution of arbitrary code in a storage system) 
%and
%\emph{software-defined storage} (the control of thin-provisioning of storage).

{\it \textbf{We define a programmable storage system}} to be a storage system that facilitates the 
re-use and extension of existing storage abstractions provided by the underlying software stack,
to enable the creation of new services via composition.
A programmable storage system can be realized by exposing existing functionality (such as 
metadata services and synchronization and monitoring capabilities) as interfaces that can be ``glued together''
in a variety of ways
using a high-level language. In contrast to programmable storage, 
\emph{active storage} differs through its focus on the injection and
execution of code within a storage system or storage device.

@@

A prototype programmable storage system that includes a non-exhaustive set of
interfaces that can be used as building blocks for constructing novel storage
abstractions, including:

@@

Building storage stacks from the ground up for a specific purpose results in
the best performance. For example, HDFS was designed specifically to serve
Hadoop's jobs, and uses techniques like exposing data locality and relaxing POSIX constraints to
to achieve application-specific I/O optimizations.~\cite{CITEME}. Alternatively, general-purpose storage
stacks are built with the flexiblity to serve many applications by providing a
variety of interfaces and tunable parameters. Unfortunately, managing competing
forces in these systems is difficult and users
want more control from the general-purpose storage stacks
without going as far as building their storage system from the ground up.

@@

The Ceph community provides empirical evidence that developers are already beginning to embrace
programmable storage. Figure~\ref{fig:obj-int-dev-growth} shows a dramatic
growth in the production use of domain-specific interfaces in the Ceph
community since 2010. In that figure, classes are functional groupings of methods on storage
objects (e.g. remotely computing and caching the checksum of an object extent).  What is most remarkable is that this trend contradicts the notion
that API changes are a burden for users.  Rather it appears that a gap in
existing interfaces are being addressed through ad-hoc approaches to programmability. In fact,
Table~\ref{table:objclasses} categorizes existing interfaces and we clearly see
a trend towards reusable services.

@@

The takeaway from Figure~\ref{fig:obj-int-dev-growth} is that programmers are
already trying to use programmability because their needs, whether they be
related to performance, availability, consistency, convenience, etc., are not
satisfied by the existing default set of interfaces. The popularity of the custom object
interface facility of Ceph could be due to a number of reasons, such as the
default algorithms/tunables of the storage system being insufficient for the
application's performance goals, programmers wanting to exploit
application-specific semantics, and/or programmers knowing how to manage resources
to improve performance. A solution based on application-specific object
interfaces is a way to work around the
traditionally rigid storage APIs because custom object interfaces give
programmers the ability to tell the storage system about their application: if
it is CPU or IO bound, if it has locality, if its size has the potential to
overload a single proxy node, etc.  The programmers know what the problem is
and how to solve it, but until the ability to modify object interfaces, they
had no way to express to the storage system how to handle their data.

Our approach is to expose more of the commonly used, code-hardened subsystems
of the underlying storage system as interfaces. The intent is that these
interfaces, which can be as simple as a redirection to the persistent data
store or as complicated as a strongly consistent directory service, should be
used and re-used in many contexts to implement a wide range of services. By
making programmability a `feature', rather than a `hack' or `workaround', we
help standardize a development process that now is largely ad-hoc.

@@

Malacology implements and demonstrates the ideas of programmable storage in
Ceph. We choose Ceph because it is a production quality system and because it
is open source. The large developer community ensures that code is
robust and the visibility of the code lets us expose any interface we want. In
the next section we describe the Ceph components that we expose as Malacology
interfaces.

@@

There are multiple reasons that make Lua an attractive runtime for implementing
Malacology. Lua is a portable, embedded scripting language that offers superior
performance and productivity trade-offs, including a JIT-based implementation
that is well known for near native performance. Additionally, Lua has been used
extensively in game engines, and systems research \cite{neto:dls14-luaos},
including storage systems where it has been effectively used both on
\cite{grawinkel:pdsw2012-lua,watkins2013:bdmc13-in-vivo,geambasu_comet_2010}
and off \cite{sevilla:sc15-mantle} the performance critical path. Finally, the
flexibility of the runtime allows execution sandboxing in order to address
security and performance concerns.

@@

In the case of Ceph, a consistent view of cluster state among server daemons
and clients is critical to provide strong consistency guarantees to clients.
Ceph maintains cluster state information in per-subsystem data structures
called ``maps'' that record
membership and status information.  A Paxos-based monitoring service is
responsible for integrating state changes into cluster maps, responding to
requests from out-of-date clients and synchronizing members of the cluster
whenever there is a change in a map so that they all observe the same system
state. As a fundamental building block of many system designs, consensus
abstractions such as Paxos are a common technique for maintaining consistent
data versions, and are a useful system to expose.

The default behavior of the monitor can be seen as a Paxos-based notification
system, similar to the one introduced in~\cite{burrows_chubby_2006}, allowing
clients to identify when new values (termed epochs in Ceph) are associated to
given maps. While Ceph does not expose this service directly, it does allow
limited use by clients through a key-value service intended for low-volume,
small size read-only configuration information. While a key-value service is
generally useful, it doesn't provide many of the useful services hidden within
the monitoring framework that would be required for applications with more
demanding requirements, such as creating arbitrary maps or associating
application-specific logic that can be executed for particular maps (or values
in a map).  Since the monitor is intended to be out of high-performance I/O
paths, a general guideline is to make use of this functionality infrequently
and to assign small values to maps.

@@

The second service implemented on Malacology is ZLog, a high-performance
distributed shared-log that is based on the CORFU
protocol~\cite{balakrishnan_corfu_2012}. A shared-log is a general, yet
powerful abstraction used to build many distributed systems, such as
cloud-based metadata management~\cite{balakrishnan:sosp13} and elastic database
storage engines~\cite{bernstein:cidr11,bernstein:vldb11,bernstein:sigmod15}.
However, existing implementations that rely on consensus algorithms such as
Paxos funnel I/O through a single point introducing a bottleneck that restricts
throughput.  In contrast, the CORFU protocol is able to achieve high throughput
using a network counter service, called a \emph{sequencer}, for obtaining the
entry at the tail of log. This metadata operation that avoids all persistent
I/O in the common case.

@@

In order to repopulate the cached tail value during recovery of a sequencer,
the maximum position in the log must be obtained. To do this, the storage
interface exposes an additional \emph{seal} method that atomically installs a
new epoch value and returns the maximum log position that has been written.
Since the recovery protocol does not inform clients of a new epoch value until
the end of the process, the validity of the tail can be guaranteed despite the
iterative process of contacting storage devices.  Recovery of the sequencer
process itself may be handled in many ways, such as leader election using an
auxiliary service such as Paxos. In our implementation, the recovery is the
same (and is inherited from) the Ceph metadata service.

@@

In Section~\ref{consistencyversioning-of-cluster-state} we described an
interface management service that is responsible for the durable and
consistent cluster-wide distribution of interface versions. We evaluate the
costs of such a service by embedding interface definitions in a Ceph cluster
membership data structure that is managed by a Paxos instance, and distributed
using a scalable, peer-to-peer gossip protocol that is built into Ceph. We
assume an architecture that consists of a light-weight interface that enforces
the semantics of interface management resulting in one or more interface
updates that must be consistent across the cluster. The following experiments
focus on the cost of achieving distribution of consistent a interface version.

@@

Figure~\ref{fig:propdelay} shows the CDF of the latency of installing and
distributing an interface update on all OSDs in the cluster. The latency is
defined as the elapsed time following the Paxos proposal for an interface
update until each OSD makes the update live (the cost of the Paxos proposal is
configurable and is discussed below). The latency measurements were taken on
the nodes running Ceph server daemons, and thus exclude the client round-trip
cost. In each of the experiments 1000 interface updates were observed.

@@

\textbf{The Malacology Approach:} Re-use Ceph's Monitor functionality to
propagate application-specific state that needs to be highly-available and
strongly consistent.  The real cost of interface management, in addition to
cluster-wide propagation of interface updates, includes the network round-trip
to the interface management service, the Paxos commit protocol, and other
factors such as system load. By default Paxos proposals occur periodically with
a 1 second interval in order to accumulate updates. In a minimum, realistic
quorum of 3 monitors using HDD-based storage, we were able to decrease this
interval to an average of 222 ms.
